---
title: "Pre_processing"
author: "COLLINS M NJAGI"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
  word_document: default
---




```{r}
#libraries

library(readxl)
library(caret)
library(tidyr)
library(dplyr)
library(corrplot)

```


```{r }
#reading data

data_GTrends <- read_excel("C:/data_science/DSE6311OM_/week3_Exploratory Data Analysis/googleTrendsMH.xlsx")

load("ACS_for_MHGoogleTrends.Rdata")

acs_data <- ACS_data
ACS_data <- NULL
```
 testing GitHub

```{r}
##CORRELATION MATRIX FOR acs_data

acs_correlation_matrix <- acs_data %>%
  select_if(is.numeric) %>%
  select(-prop_persons_below_poverty_threshold, -prop_veterans_disability) %>%
  cor()

print(acs_correlation_matrix)

```
updated code

```{r}

#presenting correlation matrix in graphic format

acs_correlation_matrix <- acs_data %>%
  select_if(is.numeric) %>%
  select(-prop_persons_below_poverty_threshold, -prop_veterans_disability) %>%
  cor() %>%
  corrplot( diag = F,
           tl.cex = 0.7,
           tl.col = "black",
           main = "acs_data correlation matrix",
           mar = c(0,0,1,0))

```

```{r}
#removing correlated features


acs_data_clean <- acs_data %>%
  select(-prop_persons_below_poverty_threshold, -prop_veterans_disability)


# convert state names into abbreviation to match state in data_GTrends

acs_data_clean$state <- toupper(state.abb[match(tolower(acs_data_clean$state), tolower(state.name))])


```
testing GitHub


```{r }
#data transformations ct variables
#creating response variable => state_mentalhealth_utili = state_psych_care / population_est

#state_mentalhealth_utili <- data_GTrends$state_psych_care / data_GTrends$population_est

data_GTrends <- data_GTrends %>%
  mutate(state_mentalhealth_util = state_psych_care/population_est,
         anxiety_prop = anxiety_ct/ population_est,
         trauma_stress_prop = trauma_stress_ct/population_est,
         adhd_prop = adhd_ct/population_est,
         bipolar_prop = bipolar_ct/population_est,
         depression_prop = depression_ct/population_est)

data_GTrends <- data_GTrends %>% 
  select(-state_psych_care, -anxiety_ct, -trauma_stress_ct, -adhd_ct, -bipolar_ct, -depression_ct)

```
  
  
```{r}
#joining both datasets acs_data and data_GTrends

GTrends_acs_joined <- inner_join(data_GTrends, acs_data_clean, by = c("year", "state"))

```


```{r }
#testing correlation

library(corrplot)

correlation_matrix <- GTrends_acs_joined %>%
  select_if(is.numeric) %>%
  select(-fips, -population_est,-private_psych_care, -total_util, -outpatient_util, -mean_anxiety, -resid_psych_care, -mean_all_trends, -mean_therapist_near_me, -state_mentalhealth_util, -trauma_stress_prop, -depression_prop, -inpatient_util, -contains(c("median", "total")),
         -total_util) %>%
  cor() 

print(correlation_matrix)

```

high correlation variables 

1. private, reside and comm_psych_care,
2.inpatient_util vs outpatient_util ( i already have state_mentalhealth_util)
3.mean_therapist near_me vs mean_psychiatrist and mean_psychologist
4.mean_alltrend vs mean_adhd, mean_ptsd, mean_anxiety, mean_mentalhospital.
5. mean_anxiety vs year, mean_adhd & ptsd
6.outpatient_util vs total_util, adhd, bipolar & depression
7.total_util
8.depression prob vs adhd. ptsd, bipolar and trauma_stress_prop
9.trauma_stress_prop vs adhd, anxiety_prop and state_mentalhealth_util
10.state_mentalhealth_util vs adhd, ptsd, bipolar

 

```{r }
#correlation matrix

GTrends_acs_joined %>%
  select_if(is.numeric) %>%
  select(-fips, -population_est,-private_psych_care, -total_util, -outpatient_util, -mean_anxiety, -resid_psych_care, -mean_all_trends, -mean_therapist_near_me, -state_mentalhealth_util, -depression_prop, -trauma_stress_prop, -inpatient_util, -contains(c("median", "total")),
         -total_util) %>%
  cor() %>%

corrplot(diag = F,
         tl.cex = 0.7,
         tl.col = "black",
         main = "Correlation Matrix of GTrends_acs_joined",
         mar = c(0, 0, 1, 0))



```

```{r}
dim(clean_GTrends_acs_joined)
```
```{r}
names(clean_GTrends_acs_joined)
```



```{r }
#data split: train and test dataset

clean_GTrends_acs_joined <- GTrends_acs_joined %>%
  select(-fips, -population_est,-private_psych_care, -total_util, -outpatient_util, -mean_anxiety, -resid_psych_care, -mean_all_trends, -mean_therapist_near_me, -depression_prop, -trauma_stress_prop, -inpatient_util, -contains(c("median", "total")), -total_util)

test_n <- (1/sqrt(20))*nrow(clean_GTrends_acs_joined)
  test_prop <- round((1/sqrt(20))*nrow(clean_GTrends_acs_joined)/nrow(clean_GTrends_acs_joined), 2)
  train_prop <- 1-test_prop
  
  print(paste0("the ideal split ratio is", train_prop, ":", test_prop, "training:testing"))
  
```


```{r}

train <- createDataPartition(clean_GTrends_acs_joined$state_mentalhealth_util,
                             p = 0.78,
                             list = FALSE,
                             times = 1)


GTrend_training_set <- clean_GTrends_acs_joined[train, ] 


test_set <- clean_GTrends_acs_joined[-train, ] 
 
dim(GTrend_training_set)
```
```{r}
dim(test_set)
```


```{r }
#data encoding

unique(clean_GTrends_acs_joined$region)
unique(clean_GTrends_acs_joined$state)
is.factor(clean_GTrends_acs_joined$region) #checking whether region is a factor =false
is.factor(clean_GTrends_acs_joined$state) #checking whether region is a factor = false


GTrend_training_set$region <- factor(GTrend_training_set$region) #convert region into a factor

     is.factor(GTrend_training_set$region) #check if its a factor to confirm = True
     levels(GTrend_training_set$region)
     class(GTrend_training_set$region)

GTrend_training_set$state <- factor(GTrend_training_set$state)

     class(GTrend_training_set$state) 
     levels(GTrend_training_set$state)


# encoding both region and state and adding them to training dataset

onehot_encoded <- dummyVars(~region + state, GTrend_training_set[,c("region", "state")],
                           levelsOnly = TRUE,
                           fullRank = TRUE)

  onehot_training <- predict(onehot_encoded, GTrend_training_set[,c("region", "state")])

  GTrend_training_set <- cbind(GTrend_training_set, onehot_training)

  GTrend_training_set_f <- GTrend_training_set %>%
  select(-region, -state) #removing the original region and state variables

#adding encoded state and region into test set dataset


test_set$region <- factor(test_set$region)
test_set$state <- factor(test_set$state)

 onehot_test <- predict(onehot_encoded, test_set[, c("region", "state")])

 test_set <- cbind(test_set, onehot_test)

 test_set_f <- test_set %>%
  select(-region, -state) 
 
 dim(GTrend_training_set_f)

```

```{r}
dim(test_set_f)
```

```{r}
head(GTrend_training_set_f[,19:26])
```


```{r }
#center and scale

test_set_f[, c(-10)] <- scale(test_set_f[, c(-10)],
                           center = apply(GTrend_training_set_f[, c(-10)], 2, mean),
                           scale = apply(GTrend_training_set_f[, c(-10)], 2, sd)) 

#(-10) is the state_mentalhealth_util, i want to exclude it from center and scale since its already a proportion. Am not sure if this is the best idea at the moment!
                                          

GTrend_training_set_f[, -10] <- scale(GTrend_training_set_f[, -10])

head(GTrend_training_set_f[,19:26])
```

```{r}
#generating codebook

library(tibble)

codebook <- tibble(
  variable = names(clean_GTrends_acs_joined),
  class = sapply(clean_GTrends_acs_joined, class),
  "Number of Missing Values" = sapply(clean_GTrends_acs_joined, function(x) sum(is.na(x))),
  "Number of Unique Values" = sapply(clean_GTrends_acs_joined, function(x) length(unique(x)))
)

print(codebook)
```




