---
title: "Pre_processing"
author: "COLLINS M NJAGI"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r reading data}

data_GTrends <- read_excel("C:/data_science/DSE6311OM_/week3_Exploratory Data Analysis/googleTrendsMH.xlsx")

```

```{r libraries}
library(caret)
library(tidyr)
library(dplyr)

```

creating response variable => state_mentalhealth_utili = state_psych_care / population_est

state_mentalhealth_utili <- data_GTrends$state_psych_care / data_GTrends$population_est
testing GitHub

```{r data transforming ct variables}


data_GTrends <- data_GTrends %>%
  mutate(state_mentalhealth_util = state_psych_care/population_est,
         anxiety_prop = anxiety_ct/ population_est,
         trauma_stress_prop = trauma_stress_ct/population_est,
         adhd_prop = adhd_ct/population_est,
         bipolar_prop = bipolar_ct/population_est,
         depression_prop = depression_ct/population_est)

data_GTrends <- data_GTrends %>% 
  select(-state_psych_care, -anxiety_ct, -trauma_stress_ct, -adhd_ct, -bipolar_ct, -depression_ct)

```


```{r testing correlation}

library(corrplot)

correlation_matrix <- data_GTrends %>%
  select_if(is.numeric) %>%
  select(-fips, -population_est,-private_psych_care, -total_util, -outpatient_util, -mean_anxiety, -resid_psych_care, -mean_all_trends, -mean_therapist_near_me, -state_mentalhealth_util, -trauma_stress_prop, -depression_prop, -inpatient_util, -contains(c("median", "total")),
         -total_util) %>%
  cor()

print(correlation_matrix)

```



 high correlation variables 
 
 1. private, reside and comm_psych_care,
 2.inpatient_util vs outpatient_util ( i already have state_mentalhealth_util)
 3.mean_therapist near_me vs mean_psychiatrist and mean_psychologist
 4.mean_alltrend vs mean_adhd, mean_ptsd, mean_anxiety, mean_mentalhospital.
 5. mean_anxiety vs year, mean_adhd & ptsd
 6.outpatient_util vs total_util, adhd, bipolar & depression
 7.total_util
 8.depression prob vs adhd. ptsd, bipolar and trauma_stress_prop
 9.trauma_stress_prop vs adhd, anxiety_prop and state_mentalhealth_util
 10.state_mentalhealth_util vs adhd, ptsd, bipolar


```{r correlation matrix}

data_GTrends %>%
  select_if(is.numeric) %>%
  select(-fips, -population_est,-private_psych_care, -total_util, -outpatient_util, -mean_anxiety, -resid_psych_care, -mean_all_trends, -mean_therapist_near_me, -state_mentalhealth_util, -depression_prop, -trauma_stress_prop, -inpatient_util, -contains(c("median", "total")),
         -total_util) %>%
  cor() %>%

corrplot(diag = F,
         tl.cex = 0.7,
         tl.col = "black",
         main = "Correlation Matrix of Curated Google Trends Dataset",
         mar = c(1, 1, 2, 1))



```


```{r data split: train and test dataset}

clean_data_GTrends <- data_GTrends %>%
  select(-fips, -population_est,-private_psych_care, -total_util, -outpatient_util, -mean_anxiety, -resid_psych_care, -mean_all_trends, -mean_therapist_near_me, -depression_prop, -trauma_stress_prop, -inpatient_util, -contains(c("median", "total")), -total_util)


train <- createDataPartition(clean_data_GTrends$state_mentalhealth_util,
                             p = 0.80,
                             list = FALSE,
                             times = 1)


GTrend_training_set <- clean_data_GTrends[train, ] 


test_set <- clean_data_GTrends[-train, ] 

```


```{r data encoding}

unique(clean_data_GTrends$region)
unique(clean_data_GTrends$state)
is.factor(clean_data_GTrends$region) 
is.factor(clean_data_GTrends$state)  


GTrend_training_set$region <- factor(GTrend_training_set$region)

     is.factor(GTrend_training_set$region) 
     levels(GTrend_training_set$region)
     class(GTrend_training_set$region)

GTrend_training_set$state <- factor(GTrend_training_set$state)

     class(GTrend_training_set$state) 
     levels(GTrend_training_set$state)

# encoding both region and state and adding them to training dataset

onehot_encoded <- dummyVars(~region + state, GTrend_training_set[,c("region", "state")],
                           levelsOnly = TRUE,
                           fullRank = TRUE)

  onehot_training <- predict(onehot_encoded, GTrend_training_set[,c("region", "state")])

  GTrend_training_set <- cbind(GTrend_training_set, onehot_training)

  GTrend_training_set_f <- GTrend_training_set %>%
  select(-region, -state)

#adding encoded state and region into test set dataset

test_set$region <- factor(test_set$region)
test_set$state <- factor(test_set$state)

 onehot_test <- predict(onehot_encoded, test_set[, c("region", "state")])

 test_set <- cbind(test_set, onehot_test)

 test_set_f <- test_set %>%
  select(-region, -state) #removing original region and state variables

```




```{r center and scale}
test_set_f[, c(-10)] <- scale(test_set_f[, c(-10)],
                           center = apply(GTrend_training_set_f[, c(-10)], 2, mean),
                           scale = apply(GTrend_training_set_f[, c(-10)], 2, sd))
                                          

GTrend_training_set_f[, -10] <- scale(GTrend_training_set_f[, -10])

```

